Testing atlantisom: true biomass output from Norwegian-Barents Atlantis

### 2019-04-30 CCAExamples.Rmd
example visualizations each model

---
title: "Testing atlantisom: true biomass output from Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'knitr', 'rmarkdown', 'tidyr', 'dplyr', 'ggplot2',
  'data.table', 'here', 'ggforce', 'ggthemes'
), 'packages.bib')
```

@@

This page documents initial testing of the atlantisom package in development at https://github.com/r4atlantis/atlantisom using three different [Atlantis](https://research.csiro.au/atlantis/) output datasets. Development of atlantisom began at the [2015 Atlantis Summit](https://research.csiro.au/atlantis/atlantis-summit/) in Honolulu, Hawaii, USA.

The purpose of atlantisom is to use existing Atlantis model output to generate input datasets for a variety of models, so that the performance of these models can be evaluated against known (simulated) ecosystem dynamics. Atlantis models can be run using different climate forcing, fishing, and other scenarios. Users of atlantisom will be able to specify fishery independent and fishery dependent sampling in space and time, as well as species-specific catchability, selectivty, and other observation processes for any Atlantis scenario. Internally consistent multispecies and ecosystem datasets with known observation error characteristics will be the atlantisom outputs, for use in individual model performance testing, comparing performance of alternative models, and performance testing of model ensembles against "true" Atlantis outputs.

Initial testing was conducted by S. Gaichas using R scripts in the R folder of this repository that are titled "PoseidonTest_[whatwastested].R". Initial tests are expanded and documented in more detail in these pages. C. Stawitz improved and streamlined the setup and intialization sections.

## Setup

First, you will want to set up libraries and install atlantisom if you haven't already. This document is written in in R Markdown [@R-rmarkdown], and we use several packages to produce the outputs [@R-tidyr; @R-dplyr; @R-ggplot2; @R-here; @R-ggforce; @R-ggthemes].

```{r message=FALSE, warning=FALSE}
library(tidyr)
require(dplyr)
library(ggplot2)
library(data.table)
library(here)
library(ggforce)
library(ggthemes)
```

If you want to load the local of the atlantisom R package, use devtools. The R package here is helpful for installing from your local atlantisom directory, as it avoids hardcoding the specific location. The code below is not running from our local atlantisom directory and is not evaluated:

```{r eval=FALSE, message=FALSE, warning=FALSE}
require(devtools)
package.dir <- here()
devtools::load_all(package.dir)
```

Or you can install directly from the Github repository.

```{r eval=FALSE, message=FALSE, warning=FALSE}
devtools::install_github("r4atlantis\atlantisom")
```

Next, load the package.

```{r message=FALSE, warning=FALSE}
library(atlantisom)
```

## Initializing input files and directories

You will first need to tell `atlantisom` to know where to look for the output and input files from your atlantis model run. Here, we will give the directory where the Atlantis inputs and outputs are stored `d.name`, the location of the functional groups file `functional.group.file` (.csv), the biomass pools file `biomass.pools.file` (.nc), the box locations file `box.file` (.bgm), an `initial.conditions.file` (.nc), the biology .prm file `biol.prm.file` (.prm), and the run .prm file `run.prm.file` (.prm). You will also need to specify a scenario name, which will be used to define the output files (i.e. output is stored in a number of netCDF files of the format: output<scenario><value>.nc). All of these files should be stored in `d.name`.

In general, Atlantis model output files that are sufficiently detailed to mimic fishery sampling in space and time are too large to store on GitHub, so are not included as examples with the atlantisom code.

In this example, we have a local folder "atlantisoutput" with three subdirectories:

  * "CalCurrentSummitScenario1": California Current Atlantis model output
  * "NEUStest20160303": older (broken) Northeast US Atlantis model output
  * "NOBACERESGlobalSustainability": Norwegian-Barents Sea Atlantis model output

Our directory structure is set up to take advantage of `here()` to allow setup on a different computer.

@@

initCCA <- FALSE
initNEUS <- FALSE
initNOBA <- TRUE

@@

This section tests the `atlantisom::create_survey()` and `atlantisom::sample_survey_biomass()` functions by comparing a census (survey sampling everything) with the results generated by `atlantisom::run_truth()` above.

To create a survey, the user specifies the timing of the survey, which species are captured, the spatial coverage of the survey, the species-specific survey efficiency ("q"), and the selectivity at age for each species.

The following settings should achieve a survey that samples all Atlantis model output timesteps, all species, and all model polygons, with perfect efficiency and full selectivity for all ages:

```{r census-spec, message=FALSE, warning=FALSE}

@@

```{r surveyBbased}

@@

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
compareB <-ggplot() +
  geom_line(data=surveyB_frombio, aes(x=time/5,y=atoutput, color="survey census B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBp4, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species. Depending on the model, we may not show all species in this example."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")

@@

```{r surveyNbased, fig.cap="Check weight at age generated from biomass_age$atoutput divided by nums$atoutput for up to 48 groups.", fig.show="hold"}

@@

@@

if(initCCA) CCAsurveyB_fromN <- surveyB_fromN
if(initNEUS) NEUSsurveyB_froN <- surveyB_fromN
if(initNOBA) NOBAsurveyB_fromN <- surveyB_fromN

@@

if(initCCA) surveyB_fromN <- CCAsurveyB_fromN
if(initNEUS) surveyB_froN <- NEUSsurveyB_fromN
if(initNOBA) surveyB_fromN <- NOBAsurveyB_fromN

#all species comparison, time intervals hardcoded for NEUS and NOBA
compareB_N <-ggplot() +
  geom_line(data=surveyB_fromN, aes(x=time/5,y=atoutput, color="survey census N->B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBfromN4, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species.Depending on the model, we may not show all species in this example."}

@@

@@

## References

>>>>

---
title: "Testing atlantisom: true biomass output from California Current Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'knitr', 'rmarkdown', 'tidyr', 'dplyr', 'ggplot2',
  'data.table', 'here', 'ggforce', 'ggthemes'
), 'packages.bib')

library(tidyr)
require(dplyr)
library(ggplot2)
library(data.table)
library(here)
library(ggforce)
library(ggthemes)
library(atlantisom)

@@

This page has visualizations for the CCA model example, Atlantis Summit Common Scenario 1. For full explanation of methods, see the file linked at the beginning of each section.

@@

initCCA <- TRUE
initNEUS <- FALSE
initNOBA <- FALSE

@@

Full methods are explained [here](https://sgaichas.github.io/poseidon-dev/TrueBioTest.html). The following settings should achieve a survey that samples all Atlantis model output timesteps, all species, and all model polygons, with perfect efficiency and full selectivity for all ages:

```{r census-spec, echo=TRUE}

@@

```{r surveyBbased, eval=FALSE}

@@

```{r readsurvey1}

if(initCCA) {
  CCAsurveyB_frombio <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBcensus.rds")))
}

if(initNEUS) {
  NEUSsurveyB_frombio <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBcensus.rds")))
}

if(initNOBA) {
  NOBAsurveyB_frombio <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBcensus.rds")))
}

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
# WHOOPS! This is CCA, really need to generalize this but hardcoded for now
compareB <-ggplot() +
  geom_line(data=surveyB_frombio, aes(x=time,y=atoutput, color="survey census B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBp4, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; next 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")

```{r matchBp5, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 5, scales="free")

@@

```{r wtagecalc, fig.cap="Check weight at age generated from biomass_age\\$atoutput divided by nums\\$atoutput for up to 64 groups.", fig.show="hold"}

@@

wtagecheck + 
  facet_wrap_paginate(~species, ncol=4, nrow=4, page=4, scales="free")

```{r surveyNbased, eval=FALSE}

@@

if(initCCA) CCAsurveyB_fromN <- surveyB_fromN
if(initNEUS) NEUSsurveyB_fromN <- surveyB_fromN
if(initNOBA) NOBAsurveyB_fromN <- surveyB_fromN

```{r readsurvey2}

if(initCCA) {
  CCAsurveyB_fromN <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBfromNcensus.rds")))
}

if(initNEUS) {
  NEUSsurveyB_fromN <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBfromNcensus.rds")))
}

if(initNOBA) {
  NOBAsurveyB_fromN <- readRDS(file.path(d.name, paste0(scenario.name, "surveyBfromNcensus.rds")))
}

@@

if(initCCA) surveyB_fromN <- CCAsurveyB_fromN
if(initNEUS) surveyB_fromN <- NEUSsurveyB_fromN
if(initNOBA) surveyB_fromN <- NOBAsurveyB_fromN

#all species comparison, time intervals hardcoded for NEUS and NOBA
# WHOOPS! Fixed for CCA but generalize this!
compareB_N <-ggplot() +
  geom_line(data=surveyB_fromN, aes(x=time,y=atoutput, color="survey census N->B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBfromN4, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; next 9 species."}

@@

```{r matchBfromN5, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species. "}

compareB_N + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 5, scales="free")

@@

### 2019-04-30 NEUSExamples.Rmd
example visualizations each model

---
title: "Testing atlantisom: true biomass output from California Current Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

@@

This page has visualizations for the CCA model example, Atlantis Summit Common Scenario 1. For full explanation of methods, see the file linked at the beginning of each section.

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
# WHOOPS! This is CCA, really need to generalize this but hardcoded for now
compareB <-ggplot() +
  geom_line(data=surveyB_frombio, aes(x=time,y=atoutput, color="survey census B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBp3, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; next 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 3, scales="free")

@@

```{r matchBp4, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; next 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")

```{r matchBp5, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 5, scales="free")

After comparing the survey census based on the biomass output of `run_truth`, we now look at a biomass index that is estimated based on numbers in the survey census and an average weight at age. This is more complex, but may be closer to the way some assessments handle survey observations.

```{r wtagecalc, fig.cap="Check weight at age generated from biomass_age\\$atoutput divided by nums\\$atoutput for up to 64 groups.", fig.show="hold"}

@@

wtagecheck + 
  facet_wrap_paginate(~species, ncol=4, nrow=4, page=3, scales="free")

wtagecheck + 
  facet_wrap_paginate(~species, ncol=4, nrow=4, page=4, scales="free")

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
# WHOOPS! Fixed for CCA but generalize this!
compareB_N <-ggplot() +
  geom_line(data=surveyB_fromN, aes(x=time,y=atoutput, color="survey census N->B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBfromN3, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; next 9 species."}

@@

```{r matchBfromN4, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; next 9 species."}

compareB_N + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")

```{r matchBfromN5, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species. "}

compareB_N + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 5, scales="free")

>>>>

---
title: "Testing atlantisom: true biomass output from Northeast US Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

@@

This page has visualizations for the NEUS model example, test03032016 (note! a not fully calibrated model). For full explanation of methods, see the file linked at the beginning of each section.

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
compareB <-ggplot() +
  geom_line(data=surveyB_frombio, aes(x=time/5,y=atoutput, color="survey census B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBp3, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 3, scales="free")

@@

After comparing the survey census based on the biomass output of `run_truth`, we now look at a biomass index that is estimated based on numbers in the survey census and an average weight at age. This is more complex, but may be closer to the way some assessments handle survey observations. 
```{r wtagecalc, fig.cap="Check weight at age generated from biomass_age\\$atoutput divided by nums\\$atoutput for up to 64 groups.", fig.show="hold"}

@@

@@

#all species comparison, time intervals hardcoded for NEUS and NOBA
compareB_N <-ggplot() +
  geom_line(data=surveyB_fromN, aes(x=time/5,y=atoutput, color="survey census N->B"), 
            alpha = 10/10) +
  geom_point(data=atBtxt2tidy, aes(x=Time/365,y=biomass, color="txt output true B"),
             alpha = 1/10) + 
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r matchBfromN3, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species."}

@@

### 2019-04-30 NOBAExamples.Rmd
example visualizations each model

---
title: "Testing atlantisom: true biomass output from Northeast US Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

@@

This page has visualizations for the NEUS model example, test03032016 (note! a not fully calibrated model). For full explanation of methods, see the file linked at the beginning of each section.

@@

```{r matchBp3, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 3, scales="free")

@@

After comparing the survey census based on the biomass output of `run_truth`, we now look at a biomass index that is estimated based on numbers in the survey census and an average weight at age. This is more complex, but may be closer to the way some assessments handle survey observations. 
```{r wtagecalc, fig.cap="Check weight at age generated from biomass_age\\$atoutput divided by nums\\$atoutput for up to 64 groups.", fig.show="hold"}

@@

```{r matchBfromN3, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species."}

@@

>>>>

---
title: "Testing atlantisom: true biomass output from Norwegian-Barents Atlantis"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

@@

This page has visualizations for the NOBA model example, CERES Global Sustainability. For full explanation of methods, see the file linked at the beginning of each section.

@@

```{r matchBp3, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; next 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 3, scales="free")

@@

```{r matchBp4, fig.cap="Testing whether the survey census gives the same results as the Atlantis output biomass index file; last 9 species."}
compareB + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")

After comparing the survey census based on the biomass output of `run_truth`, we now look at a biomass index that is estimated based on numbers in the survey census and an average weight at age. This is more complex, but may be closer to the way some assessments handle survey observations.

```{r wtagecalc, fig.cap="Check weight at age generated from biomass_age\\$atoutput divided by nums\\$atoutput for up to 48 groups.", fig.show="hold"}

@@

```{r matchBfromN3, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; next 9 species."}

@@

```{r matchBfromN4, fig.cap="Testing whether the survey census based on numbers converted to biomass gives the same results as the Atlantis output biomass index file; last 9 species. "}

compareB_N + 
  facet_wrap_paginate(~species, ncol=3, nrow = 3, page = 4, scales="free")
